Neuro V

### 2017-09-06 neuro5.rmd
trialling images, also cbf 100% completing sc1, should be rihgt

-  Generally weak
-  No sensory or paraesthesia: do exam first, but suggests against 
-  Head, neck, or shoulder pain
-  No recent injuries: not trauma
-  No pattern to weakness, not aware of stuff that makes it better or worse
-  No fatigueable element: less likely NMJ
-  No bladder or bowel disturbances, no double vision
-  No difficulties with swallowing or changes to voice/speech: not cranial nerves?
-  PMHx relatively uneventful: no post-polio

@@

@@

MND: often Dx of exclusion

-  ALS: combination of UMN and LMN in the limbs and bulbar muscles
-  Progressive muscle atrophy with only lower motor neuron signs initially; some patients develop later lower motor neurone signs
-  Primary lateral sclerosis, with predominantly upper motor neuron features
-  Progressive bulbar palsy, with degeneration of bulbar nuclei and preservation of anterior horn cell and upper motor neuron function.

@@

-  [Good DDx for weakness](http://www.dartmouth.edu/~dons/part_2/chapter_12.html)
-  [AFP article: evaluation of pt with muscle weakness](http://www.aafp.org/afp/2005/0401/p1327.html)
-  [Medscape MND](http://emedicine.medscape.com/article/2111360-overview#a1)
-  [AFP article of MND](http://www.racgp.org.au/download/documents/AFP/2011/December/201112zoing.pdf)

>>>>

-  Generally weak: something systemic, or just feeling shit
-  No sensory or paraesthesia: do exam first, but suggests against radiculopathies
-  No head, neck, or shoulder pain
-  No recent injuries: against trauma
-  No pattern to weakness, not aware of stuff that makes it better or worse: probably not mechanical/orthopaedic such as nerve entrapment
-  No fatigueable element: less likely NMJ
-  No bladder or bowel disturbances, no double vision
-  No difficulties with swallowing or changes to voice/speech: less likely cranial nerves involved
    -  CN V: muscles for mandible, tensor veli palatini
    -  CN VII: motor muscles of facial expression + stapedius, sensory ante $\frac{2}{3}$ tongue
    -  CN IX: motor middle pharyngeal constrictor + stylopharyngeus, sensory post $\frac{1}{3}$ tongue, velum, tonsils
    -  CN X: motor majority of viscera, sends motor commands to pharyngeal constrictors, cricothyroid for changes in pitch, glossopalatine and levator veli palatine muscles
    -  CN XI: trapezius and SCM of the neck
    -  CN XII: motor controlling tongue movement (intrinsic + extrinsic). Lesions here can result in fasciculations.
-  PMHx relatively uneventful: no post-polio

@@

Clin findings suggestive of MND

-  Asymm weakness in upper or lower limbs
-  Brisk reflexes in wasted limb
-  Weakness in absence of pain or prominent sensory symptoms
-  Relentless progression of symptoms and signs

@@

https://i.imgur.com/fUq88Wh.png <!-- image to be knit into the doc, under images/ -->

| Dx                    | Positives                   | Negatives                   |
|-----------------------|-----------------------------|-----------------------------|
| Peripheral neuropathy | Explains weakness           | Hyperreflexia               |
| Cervical cord lesion  |                             | Nothing below the level     |
| Muscular dystrophy    | Explains bilateral weakness | Reflexes increased, PAINFUL |
| R ulnar nerve lesion  | Weakness R hand             | Doesn't explain bilateral   |
| Motor neuron disease  | Vague as fuck               |                             |

MND: often Dx of exclusion. Generally occurs when age $\in[30,60]$, with degeneration of ante horn cells in spinal cord, motor nuclei of lower cranial nerves, and corticospinal/bulbar pathways. Usually sporadic, some familial. 5 varieties based on clinical grounds:

1.  Progressive bulbar palsy: bulbar involvement predominates owing to disease processes affecting primarily the motor nuclei of the cranial nerves
2.  Pseudobulbar Palsy: bilateral corticobulbar disease reflecting UMN dysfn. "Pseudo-bulbar effect" with uncontrollable laughing/crying to stimuli
3.  Progressive spinal muscular atrophy: LMN deficit in limbs due to degen of ante horn cells
4.  Primary lateral sclerosis: purely UMN deficit in the limbs
5.  Amyotrophic lateral sclerosis: mix of U and LMN in the limbs, sometimes associated with cognitive decline (frontotemporal dementia), psudobulbar effect, or parkinsonism

Differences between bulbar and pseudobulbar palsy (signs and symptoms):

-  Bulbar: LMN lesion to cranial nerves IX, X, XI, XII, in the medulla or lesons of the lower cranial nerves outside the brainstem. Symptoms = {tremulous lips, tongue weak and wasted and fasculations, drooling + dysphagia (unable to swallow), dysphonia, articulation issues (cant pronounce *r*)}
-  Pseudobulbar: lesion to corticobulbar tracts. Bilateral damage must be present as many muscles have bilateral innervation. Presentation = {paralysed tongue and no fasciculations, dribbling, facial muscles paralysed, exaggerated reflexes, dysphonia, dysphagia, emotional lability may be present}

Further Ix

-  Electromyography: may show changes of chronic partial denervation, ab~N spontaneous activity in resting muscle. Dx should not be made with confidence unless change are found in $\geq3# spinal regions, or $2$ regions + bulbar 
-  Motor conduction velocity is usually normal
-  Sensory conduction studies normal
-  Biopsy of wasted muscle
-  Serum CK may be elevated, but not as high as muscular dystrophies
-  CSF normal

What are possible future complications

-  Sialorrhoea: unintentional loss of saliva through the mouth
    -  Choking
    -  Pneumonia
-  Dermatological: rashes
-  Resp failure
-  Death

Principles of Mx for MND

-  Refer
-  Admit for initiation or titration of noninvasive ventilation for those with low inspiratory pressure, may prolong survival
-  Riluzole: reduces presynaptic release of Glu, increases short term survival for ALS
-  Symptomatic/supportive measures: trials of anticholinergics
    -  Braces/walker
    -  Physical therapy to prevent contractures
-  Relieve pseudobulbar affect with combination dextromethorphan/quinidine
-  Gastronomy tube if dysphagia is severe
-  Palliative

@@

-  [Good DDx for weakness](http://www.dartmouth.edu/~dons/part_2/chapter_12.html)
-  [AFP article: evaluation of pt with muscle weakness](http://www.aafp.org/afp/2005/0401/p1327.html)
-  [Medscape MND](http://emedicine.medscape.com/article/2111360-overview#a1)
-  [AFP article of MND](http://www.racgp.org.au/download/documents/AFP/2011/December/201112zoing.pdf)
-  [Categories of MND](http://accessmedicine.mhmedical.com.ezproxy.library.uq.edu.au/content.aspx?sectionid=135716257&bookid=1843&jumpsectionID=135717169&Resultclick=2)
